---
title: "VTPEH 6270 - Check Point 02"
subtitle: "Data Set Selection & R markdown Report Outline"
author: "Jiaru Wei"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Introduction

## Background 
COVID-19 testing data are a practical way to see how fast transmission is changing and where outbreaks are concentrated. I am interested in this dataset because it provides daily testing volume and percent positive at the county, regional, and statewide levels, which supports quick comparisons across places and across time. This is useful for spotting waves, comparing counties, and understanding how testing patterns may relate to changes in public health guidance and community risk.

## Project question
How did COVID-19 test positivity change over time (2020–2023) in New York State, and how did trends differ across counties and regions?

## Data set overview 
This public dataset from Health Data NY (New York State Department of Health) reports daily, aggregated counts of COVID-19 tests performed and positive results in New York State, USA. Records are organized by Test Date and Geography (62 counties, 10 regions, and statewide). Key variables include New Positives, Total Number of Tests Performed, Test % Positive, and cumulative totals. In the provided table, the time period runs from March 1, 2020 to August 30, 2023.

**Dataset link:** https://health.data.ny.gov/Health/New-York-State-Statewide-COVID-19-Testing-Archived/xdss-u53e/data_preview

# Loading data

```{r, results='hide'}

# Set working directory
setwd("D:/HuaweiMoveData/Users/111/Desktop/6270/checkpoint #02")

# Load file
data_covid_raw <- read.csv("New York State Statewide COVID-19 Testing-2023.csv")

# Preview the data
head(data_covid_raw)

```

## Common packages for this data

```{r, results='hide'}

library(janitor)
library(tidyverse)
library(ggplot2)
library(knitr)

```

# Exploring the Data

## Make variable names easy to use

```{r, results='hide'}

data_covid_raw2 <- data_covid_raw %>%
  janitor::clean_names()

```

## To make names more explicit, rename County/Geography into clearer geography fields. In this dataset, the "county" column also contains REGION and STATEWIDE labels, so renaming makes it more accurate. Perform type conversion for dates, numbers (removing commas), and percentages.

```{r, results='hide'}

data_covid <- data_covid_raw2 %>%
  rename(
    geo_name = county,
    geo_level = geography,
    cum_positives = cumulative_number_of_positives,
    tests_performed = total_number_of_tests_performed,
    cum_tests_performed = cumulative_number_of_tests_performed,
    pct_positive = test_positive
  ) %>%
  mutate(
    test_date = as.Date(test_date, format = "%m/%d/%Y"),
    across(
      c(new_positives, cum_positives, tests_performed, cum_tests_performed),
      ~ readr::parse_number(.x)
    ),
    pct_positive = readr::parse_number(pct_positive) / 100,
    geo_level = factor(geo_level, levels = c("STATEWIDE", "REGION", "COUNTY"))
  ) %>%
  arrange(test_date, geo_level, geo_name) %>%
  distinct()

glimpse(data_covid)

```

## Export the organized data into a new CSV file to facilitate subsequent analysis directly from the prepared dataset.

```{r, results='hide'}

write.csv(data_covid, "ny_covid_testing_clean.csv", row.names = FALSE)

```

## Export a variable dictionary to document any renaming and keep the workflow reproducible. Save a CSV that maps the original cleaned variable names "data_covid_raw2" to the final analysis-ready names "data_covid". Use this file "ny_covid_variable_dictionary.csv" as the reference for how each variable was labeled after cleaning.

```{r, results='hide'}

var_dictionary <- tibble::tibble(
  name_in_data_covid_raw2 = names(data_covid_raw2),
  name_in_data_covid = names(data_covid)
)

write.csv(var_dictionary, "ny_covid_variable_dictionary.csv", row.names = FALSE)

```

## Use the cleaned dataset to create a variable overview table. List each variable (max 10), assign a variable type (categorical, discrete, continuous, or date), and report its current R class (e.g., factor, numeric, Date). Include a clear caption for the table.

```{r var_overview_table}

caption_text <- paste(
  "Table 1. Variable overview for the cleaned New York State COVID-19 testing dataset",
  "(types and R classes)."
)

var_overview <- tibble::tibble(
  variable = names(data_covid),
  type = dplyr::case_when(
    variable == "test_date" ~ "date (temporal)",
    variable %in% c("geo_name", "geo_level") ~ "categorical",
    variable %in% c(
      "new_positives", "cum_positives",
      "tests_performed", "cum_tests_performed"
    ) ~ "discrete (count)",
    variable == "pct_positive" ~ "continuous (proportion)",
    TRUE ~ "other"
  ),
  class_in_R = sapply(
    data_covid,
    function(x) paste(class(x), collapse = ", ")
  )
)

knitr::kable(var_overview, caption = caption_text, booktabs = TRUE)

```

# Data Visualization

## Histogram

### Select one continuous variable and create a histogram using ggplot2. Apply a non-default theme; add clearly distinguishable vertical lines for the mean and median with a legend; include axis labels with units (use % here); and include a figure caption in the output.

```{r}

library(scales)

df_hist <- data_covid %>%
  dplyr::filter(geo_level == "STATEWIDE") %>%
  dplyr::filter(!is.na(pct_positive))

stats_lines <- tibble::tibble(
  stat = c("Mean", "Median"),
  xintercept = c(mean(df_hist$pct_positive), median(df_hist$pct_positive))
)

ggplot(df_hist, aes(x = pct_positive)) +
  geom_histogram(binwidth = 0.01) +
  geom_vline(
    data = stats_lines,
    aes(xintercept = xintercept, color = stat, linetype = stat),
    linewidth = 1
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    breaks = seq(0, 0.30, 0.05)
  ) +
  coord_cartesian(xlim = c(0, 0.30)) +
  labs(
    x = "Test positivity (%)",
    y = "Number of days",
    color = "Statistic",
    linetype = "Statistic"
  ) +
  theme_classic() +
  theme(legend.position = "top")

```

## Scatterplot

### Select two continuous variables and create a scatterplot using ggplot2. Apply a non-default theme; include axis labels with appropriate units (tests as counts, positivity as %); use relevant breaks; and include a figure caption in the output.



```{r scatter_tests_vs_positivity, fig.width=9, fig.height=5, fig.cap="Figure 2. Scatter plot of daily COVID-19 tests performed versus test positivity in New York State (STATEWIDE level). Axes are zoomed to reduce the influence of extreme values."}

df_scatter <- data_covid %>%
  dplyr::filter(geo_level == "STATEWIDE") %>%
  dplyr::filter(!is.na(tests_performed), !is.na(pct_positive))

x_max <- quantile(df_scatter$tests_performed, 0.99, na.rm = TRUE)

ggplot(df_scatter, aes(x = tests_performed, y = pct_positive)) +
  geom_point(alpha = 0.25, size = 1, color = "steelblue") +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    breaks = seq(0, 0.40, 0.05)
  ) +
  scale_x_continuous(
    labels = scales::label_comma(),
    breaks = scales::breaks_pretty(n = 6)
  ) +
  coord_cartesian(xlim = c(0, x_max)) +
  labs(
    x = "Tests performed (count per day)",
    y = "Test positivity (%)"
  ) +
  theme_classic()


```

# AI Use Disclosure Statement

*As part of this assignment, please indicate whether you used any AI-based tools (e.g., ChatGPT, Claude, Copilot, Gemini, etc.). Indicate:*

* *Did you use AI? Yes / No*

* *If yes: Write a short disclosure statement (2–3 sentences) describing:*

* *Which tool(s) you used.*

* *How you used it (e.g., to help decide on an analysis approach, to generate a first draft of code, to improve or debug code you had written yourself, etc.).*

*Example: “This document was generated using Claude to generate original code, which was then reviewed and adjusted” or "This document was generate using ChatGPT to assist with debugging of code generated by the author".*

*Please keep your statement brief and honest. The goal is transparency, not detail: we do not need exact prompts or transcripts, just a clear sense of how AI supported this work.*

This assignment used ChatGPT to help generate and format R Markdown content and R code for data cleaning, documentation, and visualization. ChatGPT was also used to troubleshoot errors and refine plots. Creating the report in R Markdown for the first time was challenging, and AI support helped structure the workflow and improve reproducibility.





