---
title: "VTPEH 6270 - Check Point 04"
subtitle: "Data Simulation"
author: "Jiaru Wei"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 1
    number_sections: true
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_knit$set(root.dir = "..")

```


# Introduction

This document serves as the foundation for the rest of the project. The final submission must meet the following criteria:

- The report must be submitted as a **PDF** file.
- It should include appropriate **section headers** and a **table of contents**.
- All **R code** used in the analysis must be included.
- All relevant **figures** must be displayed in the report.
- Console output may be included when relevant.
- The report should be clearly written, well-organized, and easy to navigate (e.g., avoid printing the full dataset within the report). ***Formatting errors may result in point deductions.***

Sections from previous reports may be reused if relevant. Any section of the report may be revised as the project develops. 
The goal of this Check Point is to gain experience with data simulation in the context of power analyses.


# Objective and Approach

## Specific Objective

Based on your data set, select one focused research question that examines the association between 2+ variables A and B (e.g., “Does X influence Y?”, “Is Y associated with Z?”, “Does Y occur more frequently in group 1 than group 2?”). The rest of the report will focus on the simulation of plausible relationships
and sampling scenarios between those variables. The ultimate goal (reached in the Final Report) is to use simulations to assess how statistical test performances, notably power (i.e., the ability to detect associations that exist) vary across biological and sampling scenarios.
Report here your research question (50 words or less).

**Is the percent of positive COVID-19 tests associated with the total number of tests conducted across New York State counties in 2023?**

## Data Description

Create a table that indicates:
- *Variable Name*
- *Variable Type in the data set* (continuous, discrete, categorical... )
- *Variable Class in R* (numeric, character, date...)

```{r}

# Load required packages
library(knitr)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales)

# Load data

data <- read_csv("data/raw/NYS_Statewide_COVID-19_Testing_2023.csv")

# Inspect structure
str(data)

# Create variable description table
variable_table <- data.frame(
  Variable_Name = c("Test Date",
                    "County",
                    "New Positives",
                    "Total Number of Tests Performed",
                    "Test % Positive"),
  
  Variable_Type_in_Data = c("Date",
                            "Categorical",
                            "Discrete (count)",
                            "Discrete (count)",
                            "Continuous (proportion)"),
  
  Variable_Class_in_R = c(class(data$`Test Date`),
                          class(data$County),
                          class(data$`New Positives`),
                          class(data$`Total Number of Tests Performed`),
                          class(data$`Test % Positive`))
)


kable(variable_table)

```

## Data Visualization

Create plots that directly address your research question using appropriate visualization types to explore the
potential association between A and B.

```{r}

# Convert variables to numeric if needed
data$`Test % Positive` <- gsub("%", "", data$`Test % Positive`)
data$`Test % Positive` <- as.numeric(data$`Test % Positive`)
data$`Total Number of Tests Performed` <- as.numeric(data$`Total Number of Tests Performed`)

# Remove missing values and zeros (log cannot handle 0)
clean_data <- data %>%
  filter(!is.na(`Total Number of Tests Performed`),
         !is.na(`Test % Positive`),
         `Total Number of Tests Performed` > 0)

# Scatter plot with log scale
ggplot(clean_data,
       aes(x = `Total Number of Tests Performed`,
           y = `Test % Positive`)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  scale_x_log10(labels = comma) +
  labs(title = "Association Between Testing Volume and Percent Positivity (Log Scale)",
       x = "Total Number of Tests Performed (log scale)",
       y = "Test % Positive") +
  theme_minimal()

```

## Description of Plausible Relationships

*Based on the visual examination of the data, describe one or several plausible relationship(s) between A and B through mathematical terms (e.g., A and B seem related through a positive linear relationship, exponential decay, threshold effect...) or literate explanations (e.g., A is higher in group 1 than group 2, A is more frequent in group A than group B...).*


Based on the scatter plot, there seems to be a negative relationship between total testing volume and percent positivity. In general, when the number of tests performed increases, the percent of positive tests tends to decrease.

The pattern does not look perfectly linear. The drop in percent positivity is much sharper when testing numbers are low, and then the relationship becomes more stable as testing increases. This suggests that the relationship may follow a log-type pattern rather than a simple straight line.

A possible way to describe this mathematically is:

$$
Y = a + b \log(X) + \epsilon
$$

where $Y$ represents percent positivity, $X$ represents the total number of tests performed, $b$ is expected to be negative, and $\epsilon$ represents random variation.

Overall, this suggests that increases in testing are associated with lower percent positivity, especially when testing levels are initially low.


## Parameters of Interest

Based on this, create a table presenting the parameters of interest, including a measure of effect size (e.g., if a linear relationship: slope $a$ and intercept $b$; if a difference in frequency between group: the probability of $Y$ occurring is $a$ times in group 1 what it is in group 2). In addition to the parameters linking $A$ to $B$, also include a noise parameter (e.g., $\epsilon$). The table should include:

- *Parameter*
- *Symbol* (e.g., $a$, $b$ and $\epsilon$)
- *Description* (e.g., unit if relevant)

*Tips* – To explore mathematical relationships between variables, you can use a graphing calculator like desmos.com.

```{r}
library(knitr)

parameter_table <- data.frame(
  Parameter = c("Intercept",
                "Effect size (slope)",
                "Noise"),
  
  Symbol = c("$a$",
             "$b$",
             "$\\epsilon$"),
  
  Description = c("Baseline percent positivity when log(X)=0",
                  "Change in percent positivity associated with a one-unit increase in log(X)",
                  "Random variability around the expected relationship")
)

kable(parameter_table, 
      format = "latex",
      booktabs = TRUE,
      caption = "Parameters of Interest for the Log-Linear Model")

View(parameter_table)

```

# Simulation

## Simulation Basis

Write a script simulating data that reflects the plausible relationship you identified between variables A and B, incorporating realistic variability and sampling scenarios. This part of the code should generate a **data frame representing a simulated data set**.

```{r}
set.seed(123)

# Define parameters
n <- 1000            # sample size
a <- 15              # intercept
b <- -0.8            # effect size (negative association)
sigma <- 3           # noise level

# Simulate testing volume (X)
# Use log-normal distribution to reflect skewed testing volumes
X <- rlnorm(n, meanlog = 8, sdlog = 1)

# Generate outcome (Y)
epsilon <- rnorm(n, mean = 0, sd = sigma)

Y <- a + b * log(X) + epsilon

# Keep percent positivity within realistic bounds (0–100)
Y[Y < 0] <- 0
Y[Y > 100] <- 100

# Create simulated dataset 
sim_data <- data.frame(
  Total_Tests = X,
  Percent_Positive = Y
)

head(sim_data)


```

## Simulation Output

Complete the simulation script with code to generate the appropriate stratified summary statistics capturing the relationship between the variables A and B (e.g., quartiles of variable A stratified by variable B). This part of the code should generate **a summary statistics data frame**.

```{r}

# Create quartile groups based on Total_Tests
sim_summary <- sim_data %>%
  mutate(Test_Group = ntile(Total_Tests, 4)) %>%
  group_by(Test_Group) %>%
  summarise(
    Q1_Tests = quantile(Total_Tests, 0.25),
    Median_Tests = median(Total_Tests),
    Q3_Tests = quantile(Total_Tests, 0.75),
    Mean_Positive = mean(Percent_Positive),
    Median_Positive = median(Percent_Positive),
    .groups = "drop"
  )

sim_summary

```

## Simulation Function

Put your simulation script into a function that returns the chosen output, and takes as parameters at least the following:
- *Effect size*
- *Noise*
- *Sample size*

```{r}

simulate_data <- function(effect_size, noise, sample_size) {

  a <- 15
  
  X <- rlnorm(sample_size, meanlog = 8, sdlog = 1)
  epsilon <- rnorm(sample_size, 0, noise)
  Y <- a + effect_size * log(X) + epsilon
  
  Y[Y < 0] <- 0
  Y[Y > 100] <- 100
  
  sim_data <- data.frame(
    Total_Tests = X,
    Percent_Positive = Y
  )
  
  sim_summary <- sim_data %>%
    mutate(Test_Group = ntile(Total_Tests, 4)) %>%
    group_by(Test_Group) %>%
    summarise(
      Mean_Positive = mean(Percent_Positive),
      .groups = "drop"
    )
  
  return(list(data = sim_data,
              summary = sim_summary))
}

```

## Simulation Automation

Run your simulation:
- For at least 10 effect sizes, 10 sample sizes, and 3 levels of noise.
- The considered sample sizes should include something the same order of magnitude as the actual sample size from your data set.
- Store the outputs in a list containing for each run of the simulation both the data and the summary statistics.

```{r}

#  Define parameter grids 

effect_sizes <- seq(-1.5, -0.1, length.out = 10)   # 10 effect sizes
noise_levels <- c(2, 3, 5)                         # 3 noise levels
sample_sizes <- c(200, 500, 1000, 2000, 5000,
                  10000, 20000, 50000, 75000, 90000)  # includes realistic size

# Initialize storage list 
simulation_results <- list()
counter <- 1

# Run simulations 
for (b in effect_sizes) {
  for (s in noise_levels) {
    for (n in sample_sizes) {
      
      sim_output <- simulate_data(
        effect_size = b,
        noise = s,
        sample_size = n
      )
      
      # Store only summary statistics (stable version)
      simulation_results[[counter]] <- list(
        effect_size = b,
        noise = s,
        sample_size = n,
        summary = sim_output$summary
      )
      
      counter <- counter + 1
    }
  }
}

# Check number of simulations
length(simulation_results)


```

# Visualization

Create a plot that visualizes your simulation results showing how the relationship between A and B varies across your tested parameters (e.g., based on the calculated summary statistics). Choose an appropriate plot format (e.g., heatmap, faceted scatter plots, or multi-panel grid) that clearly displays the key patterns. The plot should allow readers to easily compare results across different parameter combinations, hence be as simple as possible (not one panel per simulation).

```{r}
library(purrr)

# Combine all simulation summaries into one dataframe
viz_data <- map_dfr(simulation_results, function(x) {
  
  x$summary %>%
    mutate(
      effect_size = x$effect_size,
      noise = x$noise,
      sample_size = x$sample_size
    )
})

```

```{r}

library(ggplot2)

# Choose one realistic sample size for clarity
viz_subset <- viz_data %>%
  filter(sample_size == 10000)

ggplot(viz_subset,
       aes(x = Test_Group,
           y = Mean_Positive,
           color = as.factor(round(effect_size, 2)),
           group = effect_size)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ noise) +
  labs(
    title = "Simulation Results: Relationship Between Testing Volume and Percent Positivity",
    x = "Testing Volume Group (Quartiles)",
    y = "Mean Percent Positive",
    color = "Effect Size"
  ) +
  theme_minimal()

```

# Interpretation

Discuss what you learned from these simulations.

**From these simulations, I realized how much the observed relationship depends on model assumptions. The effect size clearly determined how strong the negative relationship looked, while noise affected how “clean” or scattered the pattern appeared. Even when the underlying relationship stayed the same, higher variability made it harder to see.**

**I also noticed how important sample size is. Larger samples produced smoother and more stable patterns, while smaller samples showed more fluctuation. This made me think about real-world data, where we often do not know the true effect or noise level. Overall, the simulations helped me understand that what we observe in data is shaped not only by the true relationship, but also by randomness and sampling variation.**


# AI Use Disclosure Statement

*As part of this assignment, please indicate whether you used any AI-based tools (e.g., ChatGPT, Claude, Copilot, Gemini, etc.). Indicate:*

* *Did you use AI? Yes \ No*

* *If yes: Write a short disclosure statement (2–3 sentences) describing:*

* *Which tool(s) you used.*

* *How you used it (e.g., to help decide on an analysis approach, to generate a first draft of code, to improve or debug code you had written yourself, etc.).*

*Example: “This document was generated using Claude to generate original code, which was then reviewed and adjusted” or "This document was generate using ChatGPT to assist with debugging of code generated by the author".*

*Please keep your statement brief and honest. The goal is transparency, not detail: we do not need exact prompts or transcripts, just a clear sense of how AI supported this work.*


This task utilizes ChatGPT to assist in generating and formatting R Markdown content and R code for data cleaning, documentation writing, and visualization processing. Simultaneously, ChatGPT is employed to debug errors and optimize chart effects. Gemini is also used to check grammar and spelling errors during paragraph writing.

